---
title: "EDA"
format: md
editor: visual
---
Project 2
Research Question:
How has union membership differs between gender, sector, industry, and education from 1973 to 2024?

Analysis Objective
Description:
The objective of this analysis is to summarize and visualize union membership trends over time, across sectors (private vs. public), industries (manufacturing, wholesale and trade retail, public administration), gender (female vs. male), and education (more than college vs. less than college). Time-series plots and group-wise averages will be analyzed to reveal any trends and disparities.

Analysis:
I will be quantifying the relationship between union membership and demographics (industry, sector, gender, education, year) using logistic regression.

Prediction:
Logistic regression is used to predict the probability of union membership based on education. I plan on extending the model to include gender, industry, and sector. This informs our understanding of which groups are more or less likely to be union members.

Main variables of interest:
year: Year of observation
industry: Industry category (manufacturing, wholesale/retail, public administration)
sector: Sector category (public or private)
gender: Gender of workers (male or female)
education: Highest education level (college degree or less than college degree)
perc_mem: Percentage of workers who are union members
total_mem: Total number of union members (in thousands)
total_employ: Total number of employed individuals (in thousands)

Data Description:
The data was originally collected by the U.S. Census Bureau and Bureau of Labor Statistics (BLS) through the Current Population Survey (CPS). For high earning individuals, researchers used Pareto distribution estimates. Annually on Unionstats.com, linear regression models are used to demonstrate the wage gaps between union and nonunion workers. You can find the data at https://www.unionstats.com/ under "By Sector and Demographic Group: 1973-2024". The dataset spans from 1973 to 2024.

Indicator Variables:
college: 1 if college-educated, 0 otherwise
less_college: 1 if less than college education, 0 otherwise
female: 1 if female, 0 otherwise
male: 1 if male, 0 otherwise
public: 1 if public sector, 0 otherwise
private: 1 if private sector, 0 otherwise
manu: 1 if manufacturing industry, 0 otherwise
wholesale: 1 if wholesale/retail industry, 0 otherwise
public_admin: 1 if public administration industry, 0 otherwise

## EDA
Pre-processing Data Explained:
```{r}
library(tidyverse) #Load tidyverse package

# Load and clean education data
# Read CSV for college graduates, skip the first 3 rows (as they are headers) and define column types
college <- read_csv("members_coll_plus.csv", skip = 3, 
                    col_types = list(col_double())) 
# Name columns
names(college) <- c("year", "sample_size","employment","members","covered",
                    "perc_mem","per_coverage")
# Clean percentage columns (truncate characters)
college$perc_mem <- as.numeric(substr(college$perc_mem, 1, 10))
college$per_coverage <- as.numeric(substr(college$per_coverage, 1, 10))
# Remove the last row (NA row that had copyright information)
college <- college[-52,]
# Add identifier for education level
college$education <- rep("college", 51)

# Repeat the same process for workers with less than a college degree
less_college <- read_csv("members_less_th_coll (1).csv", skip = 3,
                    col_types = list(col_double()))
names(less_college) <- c("year", "sample_size","employment","members","covered",
                         "perc_mem","per_coverage")
less_college$perc_mem <- as.numeric(substr(less_college$perc_mem, 1, 10))
less_college$per_coverage <- as.numeric(substr(less_college$per_coverage, 1, 10))
less_college <- less_college[-52,]
less_college$education <- rep("less_college", 51)

# Load and clean Gender data
# Female
female <- read_csv("members_fem (2).csv", skip = 3,
                    col_types = list(col_double()))
names(female) <- c("year", "sample_size","employment","members","covered",
                         "perc_mem","per_coverage")
female$perc_mem <- as.numeric(substr(female$perc_mem, 1, 3))
female$per_coverage <- as.numeric(substr(female$per_coverage, 1, 3))
female <- female[-52,]
female$sex <- rep("female", 51)


# Male
male <- read_csv("members_male (2).csv", skip = 3,
                    col_types = list(col_double())) 
names(male) <- c("year", "sample_size","employment","members","covered",
                "perc_mem","per_coverage")
male$perc_mem <- as.numeric(substr(male$perc_mem, 1, 3))
male$per_coverage <- as.numeric(substr(male$per_coverage, 1, 3))
male <- male[-52,]
male$sex <- rep("male", 51)

# Load and clean Sector/Industry data
# Public sector
public <- read_csv("Public.csv", skip = 2,
                    col_types = list(col_double())) 
names(public) <- c("year", "sample_size","employment","members","covered",
                   "perc_mem","per_coverage")
public$perc_mem <- as.numeric(substr(public$perc_mem, 1, 3))
public$per_coverage <- as.numeric(substr(public$per_coverage, 1, 3))
public <- public[-52,c(1:7)]
public$industry <- rep("public", 51)

# Private sector
private <- read_csv("members_priv.csv", skip = 3,
                    col_types = list(col_double())) 
names(private) <-c("year", "sample_size","employment","members","covered",
                   "perc_mem","per_coverage")
private$perc_mem <- as.numeric(substr(private$perc_mem, 1, 10))
private$per_coverage <- as.numeric(substr(private$per_coverage, 1, 10))
private <- private[-52,]
private$industry <- rep("private", 51)

# Manufacturing
manufacturing <- read_csv("members_manuf.csv", skip = 3,
                    col_types = list(col_double())) 
names(manufacturing) <- c("year", "sample_size","employment","members","covered",
                          "perc_mem","per_coverage")
manufacturing$perc_mem <- as.numeric(substr(manufacturing$perc_mem, 1, 10))
manufacturing$per_coverage <- as.numeric(substr(manufacturing$per_coverage, 1, 10))
manufacturing <- manufacturing[-52,]
manufacturing$industry <- rep("manu", 51)

# Wholesale/Retail
wholesale <- read_csv("members_whole_ret.csv", skip = 3,
                    col_types = list(col_double())) 
names(wholesale) <- c("year", "sample_size","employment","members","covered",
                      "perc_mem","per_coverage")
wholesale$perc_mem <- as.numeric(substr(wholesale$perc_mem, 1, 10))
wholesale$per_coverage <- as.numeric(substr(wholesale$per_coverage, 1, 10))
wholesale <- wholesale[-52,]
wholesale$industry <- rep("wholesale", 51)

# Public administration
public_admin <- read_csv("members_public_admin.csv", skip = 3,
                    col_types = list(col_double()))
names(public_admin) <- c("year", "sample_size","employment","members","covered",
                         "perc_mem","per_coverage")
public_admin$perc_mem <- as.numeric(substr(public_admin$perc_mem, 1, 10))
public_admin$per_coverage <- as.numeric(substr(public_admin$per_coverage, 1, 10))
public_admin <- public_admin[-52,]
public_admin$industry <- rep("public_admin", 51)

# Combine and save processed data sets

### Table for model with education: Combine college and less_college into a single education dataset
education_data <- tibble(rbind(college, less_college))
write_csv(education_data, file = "education_data.csv")

### Table for model with gender: Combine male and female into a single sex data set
sex_data <- tibble(rbind(male, female))
write_csv(sex_data, file = "sex_data.csv")

### Table for model with industry: Combine all industry/sector groups into a single industry data set
industry_data <- tibble(rbind(private, public, manufacturing,
                              wholesale, public_admin))
write_csv(industry_data, file = "industry_data.csv")

```
Load packages and read data
```{r}
#load packages and read data
library(tidyverse)

education_data <- read_csv(file="../Project 1/Data/education_data.csv")
sex_data <- read_csv(file="../Project 1/Data/sex_data.csv")
industry_data <- read_csv(file="../Project 1/Data/industry_data.csv")

```
Create Univariate, Bivariate, Multivariate, and Logistic Regression Plots
```{r}

### Model 1: Linear Regression with Education
# Predicts union membership percentage based on education level
mod1<-lm(perc_mem~education, data=education_data)
# Output summary of the model
summary(mod1)

### Model 2: Linear Regression with Gender (Sex)
# Predicts union membership percentage based on sex
mod2<-lm(perc_mem~sex, data=sex_data)
summary(mod2)

### Model 3: Linear Regression with Industry
# Predicts union membership percentage based on industry
mod3<-lm(perc_mem~industry, data=industry_data)
summary(mod3)

### Load libraries
library(ggplot2) # For data visualization
library(dplyr)   # For data manipulation (if needed)
library(Hmisc)   # For mean_cl_boot confidence intervals

### Table 1: Boxplot of Union Membership % by Education Level
plot1 <- ggplot(education_data, aes(x = education, y = perc_mem, fill = education)) +
  geom_boxplot(alpha = 0.5) + # Semi-transparent boxplot
  stat_summary(fun = mean, geom = "point", shape = 20, size = 4, color = "black") +
  labs(title = "Union Membership % by Education",
       x = "Education Level", y = "Union Membership %") +
  theme_minimal() 

ggsave("table1_education.png", plot = plot1, width = 6, height = 4, dpi = 300)
print(plot1) # Display plot

### Table 2: Boxplot of Union Membership % by Sex
plot2 <- ggplot(sex_data, aes(x = sex, y = perc_mem, fill = sex)) +
  geom_boxplot(alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 4, color = "black") +
  labs(title = "Union Membership % by Gender",
       x = "Sex", y = "Union Membership %") +
  theme_minimal()

ggsave("table2_sex.png", plot = plot2, width = 6, height = 4, dpi = 300)
print(plot2)

### Table 3: Boxplot of Union Membership % by Industry
plot3 <- ggplot(industry_data, aes(x = industry, y = perc_mem, fill = industry)) +
  geom_boxplot(alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 20, size = 4, color = "black") +
  labs(title = "Union Membership % by Industry",
       x = "Industry", y = "Union Membership %") +
  theme_minimal()

ggsave("table3_industry.png", plot = plot3, width = 7, height = 4, dpi = 300)
print(plot3)

### Table 4: Logistic Regression - Predicted Probability by Education
set.seed(123)  # For reproducibility
n <- 100  # Sample size for simulation
education <- rbinom(n, 1, 0.5)  # Simulate education variable (0 = <College, 1 = College)

# Generate probabilities using a logistic model
prob_union <- 1 / (1 + exp(-( -1 + 2 * education )))  

# Simulate union membership outcome based on probabilities
union_member <- rbinom(n, 1, prob_union)

# Create data frame for modeling
df2 <- data.frame(union_member, education)

# Fit logistic regression model
model2 <- glm(union_member ~ education, data = df2, family = binomial)
df2$predicted_prob <- predict(model2, type = "response")  # Get predicted probabilities

# Plot predicted probabilities by education level
plot5 <- ggplot(df2, aes(x = factor(education), y = predicted_prob)) +
  stat_summary(fun = mean, geom = "point", size = 5, color = "blue") +  # Show mean probabilities
  stat_summary(fun.data = mean_cl_boot, geom = "errorbar", width = 0.1, color = "darkblue") + # Add error bars
  labs(
    x = "Education (0 = <College, 1 = College)",
    y = "Predicted Probability of Union Membership",
    title = "Logistic Regression: Probability by Education"
  ) +
  theme_minimal()

ggsave("table5_logistic_education.png", plot = plot5, width = 6, height = 4, dpi = 300)
print(plot5)

### Table 5: Multivariate Plot - Union Membership % by Industry and Sex
set.seed(123)  # For reproducibility
industry_data$sex <- sample(c("male", "female"), size = nrow(industry_data), replace = TRUE)  # Simulate sex variable

# Create multivariate boxplot
plot_multi <- ggplot(industry_data, aes(x = industry, y = perc_mem, fill = sex)) +
  geom_boxplot(alpha = 0.5, position = position_dodge(width = 0.8)) + # Separate boxes by sex
  stat_summary(fun = mean, geom = "point", aes(group = sex), 
                position = position_dodge(width = 0.8), shape = 20, size = 3, color = "black") +
  labs(title = "Union Membership % by Industry and Sex",
       x = "Industry", 
       y = "Union Membership %") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") # Use color palette for better visualization

ggsave("table4_multivariate.png", plot = plot_multi, width = 8, height = 5, dpi = 300)
print(plot_multi)

### Table 6: Univariate Plot - Histogram of Union Membership %
# Visualize distribution of union membership percentages
plot_uni <- ggplot(industry_data, aes(x = perc_mem)) +
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black", alpha = 0.7) + # Histogram settings
  labs(title = "Distribution of Union Membership %",
       x = "Union Membership %",
       y = "Frequency") +
  theme_minimal()

print(plot_uni)
ggsave("table5_univariate_histogram.png", plot = plot_uni, width = 6, height = 4, dpi = 300)
```
